### 2020-05-15 intro.tex
add missing images
cleaned up contribution summary

Our focus in this paper is on addressing a number of non-trivial technical 
deficiencies that limit both palette providers and palette clients in Graphite. 
The system of \emph{live literals}, or \emph{livelits}, introduced in
this work and demonstrated in Fig.~\ref{fig:color}(b), differs from Graphite  
in that it achieves each of the following goals.
\begin{enumerate}
  \setlength\itemsep{0.5em}
  \item \textbf{Decentralized Extensibility}: 
    In Graphite, there can be at most one palette per type, 
    and it must be associated with that type at its definition site, i.e. by the provider of that type.
    By contrast, livelit definitions stand apart from type definitions 
    and clients invoke livelits explicitly by name (names are prefixed with the \li{\$} symbol, 
    e.g. \li{\$color} and \li{\$slider} in Fig.~\ref{fig:color}(b), to distinguish
    livelit names from standard variables), so these limitations do not apply.
    We call this \emph{decentralized extensibility}.
  \item \textbf{Persistence}: Graphite palettes are {ephemeral}, 
  i.e. they disappear once the initial interaction is complete, 
  leaving behind only the generated textual code.\footnote{Graphite does include an \emph{ad hoc} mechanism that allows palettes to parse the code that is selected in the editor 
  when the palette loads, but this requires that each palette implement a parser for the subset of Java used in the code that it generates,
  and therefore this mechanism is quite brittle. It is also difficult to persist GUI state that is not included in the generated code.}
  This means that only the programmer that initially enters the expression 
  benefits from the feedback and affordances provided by the palette. 
  By contrast, livelits are persistent. We use a pure functional model-view-update architecture 
  (i.e. the Elm architecture\todo{what to cite?}{}) as our GUI framework, 
  so only the model needs to be persisted, rather than the full GUI state.  
  We use the word ``literal'' rather than ``palette'' because, with support for persistence, livelits
  are more analogous to literal notation, e.g. list literals, than palettes as they exist in graphical end-user 
  applications.

  \item \textbf{Compositionality}: 
  Graphite does not provide any way to {enter subexpressions within a palette}.
  At best, they can contain text boxes, but expressions entered into these text boxes 
  do not have any editor support (which creates many problems, one of which is that they cannot themselves be generated by palettes). 
  This implies that palettes can only reasonably generate {closed expressions}. 
  For example, in Figure~\ref{fig:color}(a), there is no way to specify that a color component 
  should be computed by a specified expression---only literal numbers are supported, so the color palette
  is useful only for generating simple color constants.
  By contrast, livelits allow subexpressions, which we call \emph{splices} following
  the terminology of \citet{TLMs} who developed reasoning principles for splicing in a purely textual setting.
  Fig.~\ref{fig:color}(b) demonstrates splicing: the client defines a variable, \li{grayLevel}, 
  for each of the red, green, and blue components, and places a slider inline in the splice governing 
  the alpha (A) component. In Sec.~X\todo{section}{}, we will see more sophisticated examples.
  \item \textbf{Liveness}: Graphite's palettes can only provide static feedback, because they do not have 
  access to run-time information. By contrast, livelits can request the evaluation result for splices,
  using it to provide live feedback, i.e. feedback that takes into account the run-time environments 
  around the livelit. For example, in Fig.~\ref{fig:color}, the color displayed requires evaluating the RGBA 
  component splices to numeric values. We will see more sophisticated examples later in the paper, including examples
  that appear inside functions.
  \item \textbf{Parameterization}: 
    \cyrus{... this would be really nice to include + demonstrate with slider bounds and later data flow + 
    we need it to do anything non-trivial with hygiene 
    ...}
\end{enumerate}
\clearpage

@@

\section{Livelit Implementation}
Which livelit implementation do we want to show? Maybe grade cutoffs? Slider as simple intro + color chooser?

Talk about each component of a livelit definition:
\begin{itemize}
  \item livelit name (talk about binding structure)
  \item expansion type
  \item model type
  \item action type (mention integration into Hazel's undo system)
  \item initial model
  \item update function (monadic actions; talk about color palette update; SpliceRefs)
  \item view function (monadic actions; talk about GUI stuff? tab order? dimensions? mostly defer to Elm for details)
  \item expand function (should we call it fill?)
\end{itemize}

@@

\section{Discussion and Conclusion}
Future work:
\begin{itemize}
  \item pattern matching
  \item type splices
  \item module splices
  \item explicitly make bindings available in splices
  \item better UI for closure provenance / connect with control flow / call stack better
  \item deriving livelits from type definitions
  \item bidirectional evaluation ala SnS
  \item integration with structure editing more cleanly
  \item \dots
  \item collaborative editing?
  \item side effects?
  \item big vision: authoring environment based on livelits
\end{itemize}
\clearpage
\appendix
\section{Misc. Material from Prior Draft}

>>>>

Our focus in this paper is on addressing several non-trivial technical 
deficiencies that limit both palette providers and palette clients in Graphite. 
The system of \emph{live literals}, or \emph{livelits}, introduced in
this paper and demonstrated in Fig.~\ref{fig:color}(b), differs from Graphite  
in each of the following ways.
\begin{enumerate}
  \setlength\itemsep{0.5em}
  \item \textbf{Decentralized Extensibility}: 
    Graphite allows only the provider of a type to define a corresponding palette. 
    By contrast, livelit definitions are orthogonal to type definitions. 
    Clients invoke livelits explicitly by name (names are prefixed by the \li{\$} symbol, 
    e.g. \li{\$color} and \li{\$slider} in Fig.~\ref{fig:color}(b)).
    We call this \emph{decentralized extensibility}.
  \item \textbf{Persistence}: Graphite's palettes are {ephemeral}, 
  i.e. they disappear after the initial interaction, 
  leaving behind only the generated textual code.%
  \footnote{Graphite does include an \emph{ad hoc} mechanism that 
  allows palettes to parse the code that is selected in the editor 
  when the palette loads, but this requires that each palette implement 
  a parser for the subset of Java used in the code that it generates,
  and therefore this mechanism is quite brittle. It is also difficult 
  to persist GUI state that is not included in the generated code.}
  Consequently, only the programmer that initially enters the expression 
  benefits from the feedback and affordances that the palette provides.
  By contrast, livelits are persistent. We use a pure functional model-view-update architecture 
  (i.e. the Elm architecture\todo{what to cite?}{}) as our GUI framework, 
  so only the model needs to be persisted, rather than the full GUI state.  
  We use the word ``literal'' rather than ``palette'' because, with persistence, livelits
  are more analogous to literal notation, e.g. list literals.

  \item \textbf{Compositionality}: 
  Graphite does not provide any way to {enter subexpressions within a palette}.
  At best, palettes can contain text boxes, but expressions entered into these text boxes 
  do not have any editor support (which creates many problems, one of which is that they cannot themselves be generated by palettes). 
  This implies that palettes can only reasonably generate {closed expressions}. 
  For example, in Figure~\ref{fig:color}(a), the color palette
  is useful only for generating simple color constants. 
  By contrast, livelits have first-class support for subexpressions, which we call \emph{splices} (following
  the terminology of \citet{TLMs} who developed reasoning principles for splicing in a purely textual setting).
  Fig.~\ref{fig:color}(b) demonstrates splicing: the client is able to define a variable, \li{grayLevel}, 
  to capture the relationship between the red, green, and blue components.
  The client is also able to use a slider inline in the splice governing 
  the alpha (A) component.
  \item \textbf{Liveness}: Graphite's palettes do not have 
  access to the run-time environment. By contrast, livelits can evaluate splices
  and therefore provide feedback related to the run-time behavior of the expression being generated. 
  For example, in Fig.~\ref{fig:color}(b), the color that is displayed is determined by evaluating the RGBA 
  component splices to numeric values.
  Evaluation occurs in a run-time environment (i.e. closure) determined by 
  leaving the hole being filled by the livelit temporarily unfilled and then evaluating
  using a two-phased variation on the semantics for incomplete programs developed by \citet{HazelnutLive}. 
  We will see more sophisticated examples later in the paper, including a livelit 
  that appears inside a function that is called multiple times, leading to multiple closures for the client 
  select from.
  \item \textbf{Parameterization}: 
    \cyrus{... this would be really nice to include + demonstrate with slider bounds and later data flow + 
    we need it to do anything non-trivial with hygiene 
    ...}
\end{enumerate}
\clearpage

@@

\section{Livelit Definitions}
Which livelit implementation do we want to show? Maybe grade cutoffs? Slider as simple intro + color chooser?

Talk about each component of a livelit definition:
\begin{itemize}
  \item livelit name (talk about binding structure)
  \item expansion type
  \item model type (must be serializable)
  \item action type (mention integration into Hazel's undo system)
  \item initial model
  \item update function (monadic actions; talk about color palette update; SpliceRefs)
  \item view function (monadic actions; talk about GUI stuff? tab order? dimensions? mostly defer to Elm for details)
  \item expand function (should we call it fill?)
\end{itemize}

@@

\section{Discussion and Conclusion}
Future work:
\begin{itemize}
  \item pattern matching
  \item type splices
  \item module splices
  \item explicitly make bindings available in splices
  \item better UI for closure provenance / connect with control flow / call stack better
  \item deriving livelits from type definitions
  \item bidirectional evaluation ala SnS
  \item integration with structure editing more cleanly
  \item \dots
  \item collaborative editing?
  \item side effects?
  \item full screening livelits (or compositions thereof) as a way to create end-user workflows
  \item big vision: authoring environment based on livelits
\end{itemize}
\clearpage
\appendix
\section{Misc. Material from Prior Draft}
